---
title: "Differential protein expression"
author: "Laura Abram adapted from Frank Stein (EMBL, PCF)"
---
This analysis pipeline is partly based on an analysis workflow developed by Bernd Klaus.

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, autodep = TRUE)
```
 
## Defining a working directory
This directory should contain the metadata file and the tab delimited text files containing the data.
```{r echo = FALSE}
setwd("C:/Users/laura/OneDrive - Cardiff University/Documents/9 - Publications/Proteomics paper/proteomics manuscript and figures V8 LA/CTX and INT annotated script/plots")
```
 
## Loading packages
```{r LoadingPackages}
library(vsn)
library(limma)
library(MSnbase)
library(gplots)
library(fdrtool)
library(biobroom)
library(tidyverse)
library(UpSetR)
library(glue)
library(rstatix)
library(ggpubr)
library(Biostrings)
library(camprotR)
library(httr)
library(openxlsx)
library(tidytext)
library(clusterProfiler)
library(org.Hs.eg.db)
library(tidyr)
library(dplyr)
```

## Defining parameters
```{r DefiningParameters}
set.seed(1234) # set seed for reproducibility

script.version <- "V1"
dir_save <- paste0(script.version, "/", Sys.Date()) #output folder for files and images
if (!dir.exists(dir_save))
{dir.create(dir_save)}

```

# Loading data and annotating experimental conditions
```{r loadRawFiles}
conditions <- read.delim("metadata_merged.csv", sep = ";") #metadata with LC-MS channels and experimental conditions
files <- file.path(unique(conditions$file))
files <- files[file.exists(files)]
data <- NULL
for (i in seq_along(files))
{
  print(files[i])
  x <- read_delim(file.path(files[i]), delim = "	")
  names(x) <- make.names(names(x))
  x$Gene[is.na(x$Gene)] <- x$Protein.ID[is.na(x$Gene)]
  
  x <- x %>%
    dplyr::filter(Unique.Peptides >= 2)%>% #keep only proteins with at least two unique peptide matches
    dplyr::filter(Razor.Peptides >= 2) %>% #keep only proteins with at least two razor peptide matches
    dplyr::filter(!grepl("rev_", Protein))%>%  #remove reverse hits
    dplyr::filter(!grepl("contam_", Protein)) #remove known contaminants
  
  x$file <- files[i]
  keepnames <- as.character(subset(conditions, file == files[i])$col.name)
  x <- x[, c("Gene", "Protein.ID", "Protein.Description", "Organism", "Unique.Peptides", "Razor.Peptides", "Total.Intensity", "file", keepnames)]
  x <- x %>%
    group_by(Gene, Protein.ID, Protein.Description, Organism, Unique.Peptides, Razor.Peptides, Total.Intensity, file) %>%
    gather(key = "col.name", value = "value", keepnames)
  x$Total.Intensity <- as.numeric(as.character(x$Total.Intensity))
  data <- bind_rows(data, x)
  rm(x, keepnames)
}
rm(i, files)
data <- left_join(data, conditions)
data <- subset(data, value > 0)
```
#cRAP

```{r cRAP}
# download the CCP cRAP file if not already present
outfile <- "CCP_cRAP.fasta"
if (!file.exists(outfile)) {
  download_ccp_crap(file = outfile, is_crap = TRUE, overwrite = TRUE, verbose = TRUE)
} 
crap_ccp <- readAAStringSet("CCP_cRAP.fasta")
crap_ids <- str_match(names(crap_ccp), "^[^|]+\\|[^|]+\\|([^|]+)")[,2] # Protein IDs from cRAP

removed_CRAP <- data %>% # Proteins that will be removed
  filter(Protein.ID %in% crap_ids)%>%
  dplyr::ungroup()%>%
  dplyr::select(c(Gene, Protein.ID, Protein.Description))
removed_CRAP <- removed_CRAP[!duplicated(removed_CRAP), ]

data <- data %>% #remove contaminants from dataset
  filter(!Protein.ID %in% crap_ids)%>%
  dplyr::ungroup()

```

#Identification overview
```{r IdentificationOverview}
data %>%
  ungroup() %>%
  dplyr::select(Gene, condition, sample, rep) %>%
  distinct() %>%
  group_split(sample) %>%  # split data into a list by sample
  purrr::walk(function(df_sample) {
    sample_name <- unique(df_sample$sample)

    #bar plot of # of identified protein per sample (condition and replicate)
    p <- ggplot(df_sample, aes(condition, fill = rep)) +
      geom_bar(position = position_dodge()) +
      scale_fill_manual(values = c("#009E73", "#5DBB9F", "#B3E2CD"), labels = c("1", "2", "3")) +
      scale_colour_manual(values = c("#009E73", "#5DBB9F", "#B3E2CD")) +
      ggtitle(paste0("Protein identification overview - Sample: ", sample_name)) +
      labs(caption = "proteins with at least 2 razor peptides and 2 unique peptides",
        x = "Sample", y = "Protein Count", fill = "Replicate")+
      theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
      theme_classic()+
      geom_text(aes(label = after_stat(count), y = after_stat(count)),
        stat = "count", size = 2, position = position_dodge(width = 0.9),
        hjust = 1, angle = 90)

    ggsave(filename = file.path(dir_save, paste0("Identification_CountOverview_",  sample_name, "_", script.version, "_", Sys.Date(), ".png")),
      plot = p, width = 5, height = 4, dpi = 600)
})

# Upset plot for overlapping identified proteins
upsetR.data <- data %>%
  mutate(Gene = paste(Gene, Protein.ID, sep = "_")) %>%
  group_by(Gene, ID) %>%
  summarise(value = sum(value, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(identified = 1) %>%
  dplyr::select(Gene, ID, identified) %>%
  spread(key = ID, value = identified, fill = 0) %>%
  as.data.frame()
colnames(upsetR.data) <- colnames(upsetR.data) %>%
  str_replace("_(?=[^_]*$)", ":")
upset.plot <- upset(upsetR.data, order.by = "freq", nsets = 24, mb.ratio = c(0.4, 0.6), nintersects = 30)
png(file = file.path(dir_save, paste0("Identification_UpSet_plot_", script.version,"_", Sys.Date(), ".png")), width = 3000, height = 3000, res=600)
print(upset.plot)
dev.off()
rm(upset.plot)
write.csv(upsetR.data, file = file.path(dir_save, paste0("Identification_Overview_", script.version, "_", Sys.Date(),  ".csv")), row.names = FALSE)

```

# Identification and removal of duplicated gene names
```{r renameDuplicates}
dup.data <- data %>%
  ungroup() %>%
  dplyr::select(Gene, Protein.ID, file, Razor.Peptides) %>%
  distinct() %>%
  group_by(file, Gene) %>%
  dplyr::filter(n() > 1) #if a gene appears in the file more than once assign to dup.data
if (nrow(dup.data) > 0) #if dup.data exists, save
{write.csv(dup.data, file.path(dir_save, paste0("Duplicates_", script.version, ".csv")), row.names = FALSE)}
dups <- unique(dup.data$Gene)
if (length(dups) > 0)
{ print("Duplicates were identified:")
  print(dups)
  print("Removing genes with smaller numbers of quantified peptides.")
  for (dup in dups){
    for (filen in unique(subset(dup.data, Gene == dup)$file)){
      sub <- subset(data, Gene == dup & file == filen)
      data <- subset(data, paste(Gene, file) != paste(dup, filen))
      sub <- subset(sub, Razor.Peptides == max(sub$Razor.Peptides, na.rm = TRUE)) #if duplicates exist, select one with the most peptides
      data <- bind_rows(data, sub)
      rm(sub)
    }
    rm(filen)
  }
  rm(dup)
}

dup.data <- data %>%
  ungroup() %>%
  dplyr::select(Gene, Protein.ID, file, Razor.Peptides) %>%
  distinct() %>%
  group_by(file, Gene) %>%
  dplyr::filter(n() > 1)
dups <- unique(dup.data$Gene)
if (length(dups) > 0)
{print("Duplicates:")
  print(dups)
  print("Removing genes with the least overlap of protein ids in other ms experiments")
  for (dup in dups){
    sub <- subset(data, Gene %in% dup)
    IDs <- unlist(strsplit(sub$Protein.ID, split = "[|]"))
    rm(sub)
    for (filen in unique(subset(dup.data, Gene == dup)$file)){
      sub <- subset(data, Gene == dup & file == filen)
      data <- subset(data, paste(Gene, file) != paste(dup, filen))
      sub$ID.count <- 0
      for (ID in unique(sub$Protein.ID)){
        sub$ID.count[sub$Protein.ID == ID] <- length(which(IDs %in% unlist(strsplit(ID, split = "[|]"))))
      }
      sub <- subset(sub, ID.count == max(sub$ID.count, na.rm = TRUE))
      sub$ID.count <- NULL
      data <- bind_rows(data, sub)
      rm(sub)
    }
    rm(filen, IDs)
  }
  rm(dup)
}
rm(dups, dup.data)
```

# Transforming long data to wide data
1 row per protein across all samples
```{r transformLongData2WideData, fig.height = 5}
#plot of log-transformed reporter ion intensity per sample to see spread and distribution
ggplot(data = data, aes(condition, log2(value), fill = rep)) + 
  geom_boxplot() + 
  ylab("log2(reporter_intensity)") + 
  facet_grid(. ~ sample, scale = "free_x", space = "free") + 
  scale_fill_manual(values = c("#009E73", "#5DBB9F", "#B3E2CD"), labels = c("1", "2", "3")  
                        ) +
      scale_colour_manual(values = c("#009E73", "#5DBB9F", "#B3E2CD")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))

cdata <- data %>%
  group_by(Gene) %>%
  mutate(key = paste("reporter_intensity", sample, condition, rep, sep = "_")) %>%
  dplyr::select(Gene, key, value) %>%
  group_by(Gene, key) %>%
  summarise(value = sum(value, na.rm = TRUE)) %>%
  spread(key = key, value = value)

fdata <- data %>%
  group_by(Gene) %>%
  dplyr::select(Gene, file) %>%
  distinct() %>%
  summarise(found.in.files = n()) #how many files each protein appears in
fdata_i <- data %>%
  mutate(file2 = make.names(file)) %>%
  group_by(Gene, file2) %>%
  dplyr::select(Gene, file, file2) %>%
  distinct() %>%
  summarise(found.in.file = n()) %>%
  mutate(file2 = paste0("found.in.file_", file2)) %>%
  spread(key = file2, value = found.in.file)
fdata <- left_join(fdata, fdata_i)
fdata_i <- data %>%
  group_by(Gene, sample) %>%
  dplyr::select(Gene, file, sample) %>%
  distinct() %>%
  summarise(found.in.files = n()) %>%
  mutate(sample = paste0("found.in.files_", sample)) %>%
  spread(key = sample, value = found.in.files)
fdata <- left_join(fdata, fdata_i)

fdata_i <- data %>%
  group_by(Gene) %>%
  dplyr::select(Gene, condition) %>%
  distinct() %>%
  summarise(found.in.conditions = n()) #how many samples/conditions each protein appears in 
fdata <- left_join(fdata, fdata_i)
fdata_i <- data %>%
  group_by(Gene, sample) %>%
  dplyr::select(Gene, condition, sample) %>%
  distinct() %>%
  summarise(found.in.conditions = n()) %>%
  mutate(sample = paste0("found.in.conditions_", sample)) %>%
  spread(key = sample, value = found.in.conditions)
fdata <- left_join(fdata, fdata_i)

fdata_i <- data %>%
  group_by(Gene) %>%
  dplyr::select(Gene, rep) %>%
  distinct() %>%
  summarise(found.in.reps = n()) #how many replicates each protein appears in 
fdata <- left_join(fdata, fdata_i)
fdata_i <- data %>%
  group_by(Gene, sample) %>%
  dplyr::select(Gene, rep, sample) %>%
  distinct() %>%
  summarise(found.in.reps = n()) %>%
  mutate(sample = paste0("found.in.reps_", sample)) %>%
  spread(key = sample, value = found.in.reps)
fdata <- left_join(fdata, fdata_i)

fdata_i <- data %>%
  group_by(Gene) %>%
  dplyr::select(Gene, Unique.Peptides) %>%
  distinct() %>%
  summarise(max.Unique.Peptides = max(Unique.Peptides, na.rm = TRUE)) #calculate number of unique peptides
fdata <- left_join(fdata, fdata_i)
fdata_i <- data %>%
  group_by(Gene, sample) %>%
  dplyr::select(Gene, sample, Unique.Peptides) %>%
  distinct() %>%
  summarise(max.Unique.Peptides = max(Unique.Peptides, na.rm = TRUE)) %>%
  mutate(sample = paste0("max.Unique.Peptides_", sample)) %>%
  spread(key = sample, value = max.Unique.Peptides)
fdata <- left_join(fdata, fdata_i)

fdata_i <- data %>%
  group_by(Gene) %>%
  dplyr::select(Gene, Razor.Peptides) %>%
  distinct() %>%
  summarise(max.Razor.Peptides = max(Razor.Peptides, na.rm = TRUE)) #calculate number of razor peptides per protein
fdata <- left_join(fdata, fdata_i)
fdata_i <- data %>%
  group_by(Gene, sample) %>%
  dplyr::select(Gene, sample, Razor.Peptides) %>%
  distinct() %>%
  summarise(max.Razor.Peptides = max(Razor.Peptides, na.rm = TRUE)) %>%
  mutate(sample = paste0("max.Razor.Peptides_", sample)) %>%
  spread(key = sample, value = max.Razor.Peptides)
fdata <- left_join(fdata, fdata_i)

#abundance overview
fdata_i <- data %>%
  group_by(Gene) %>%
  dplyr::select(Gene, Total.Intensity) %>%
  distinct() %>%
  summarise(average.Total.Intensity = mean(Total.Intensity, na.rm = TRUE)) #calculate mean total reporter ion intensity across all samples
fdata <- left_join(fdata, fdata_i)
fdata_i <- data %>%
  group_by(Gene, sample) %>%
  dplyr::select(Gene, sample, Total.Intensity) %>%
  distinct() %>%
  summarise(average.Total.Intensity = mean(Total.Intensity, na.rm = TRUE)) %>%
  mutate(sample = paste0("average.Total.Intensity_", sample)) %>%
  spread(key = sample, value = average.Total.Intensity)
fdata <- left_join(fdata, fdata_i)
id_data <- data %>%
  ungroup() %>%
  dplyr::select(Gene, Protein.ID, Protein.Description, Organism) %>%
  distinct()
dups <- id_data %>%
  group_by(Gene) %>%
  summarize(n = n()) %>%
  dplyr::filter(n > 1) %>%
  dplyr::select(Gene)
if (nrow(dups) > 0)
{ dups <- dups$Gene
  dup.data <- subset(id_data, Gene %in% dups)
  id_data <- subset(id_data, !Gene %in% dups)
  combine_ids <- function(ids)
  { ids <- as.character(ids)
    return(paste(
      sort(unique(unlist(strsplit(ids, split = "[;]")))), collapse = ";"))
  }
  dup.data <- dup.data %>%
    group_by(Gene) %>%
    summarise(Protein.ID = combine_ids(Protein.ID), 
              Protein.Description = combine_ids(Protein.Description), 
              Organism = combine_ids(Organism))
  id_data <- bind_rows(id_data, dup.data)
  rm(dup.data, combine_ids)
}
rm(dups)
fdata <- left_join(id_data, fdata)
rm(id_data)
cdata <- left_join(fdata, cdata)
rm(fdata_i, fdata)
```

# Filter data
At least 2 razor peptides and 2 unique peptides for a protein to be kept in the dataset
```{r filterData}
dim(cdata)
min.found.in.files <- 1
min.Razor.Peptides <- 2

for (s in unique(conditions$sample))
{ cname <- paste0("keep_", s)
  cdata[, cname] <- ifelse(
    cdata[, paste0("found.in.files_", s)] >= 
      min.found.in.files & cdata[, paste0("max.Razor.Peptides_", s)] >= min.Razor.Peptides, TRUE, FALSE)
  cdata[is.na(cdata[, cname]), cname] <- FALSE
  rm(cname)}
rm(s)
cdata$keep <- FALSE

min.Unique.Peptides <- 2
for (s in unique(conditions$sample))
{ cname <- paste0("keep_", s)
  cdata[, cname] <- ifelse(
    cdata[, paste0("found.in.files_", s)] >= 
      min.found.in.files & cdata[, paste0("max.Unique.Peptides_", s)] >= min.Unique.Peptides, TRUE, FALSE)
  cdata[is.na(cdata[, cname]), cname] <- FALSE
  rm(cname)}
rm(s)
cdata$keep <- FALSE
for (i in seq_along(rownames(cdata)))
{ cdata$keep[i] <- any(as.logical(cdata[i, grep("keep_", names(cdata), value = TRUE)]))}
rm(i)
cdata <- subset(cdata, keep)
dim(cdata)
rm(min.found.in.files, min.Razor.Peptides, min.Unique.Peptides)

#summarise number of proteins kept and number removed
n_unique<-data %>%
  group_by(sample) %>%
  summarise(n_unique_proteins = n_distinct(Protein.ID))
print(paste0("# unique proteins after all filtering:", n_unique))

#save long-format data after contaminants, reverse hits and <2 unique/razor peptides removed
write.csv(data, file.path(dir_save, paste0("Data after quality control_", script.version, "_", Sys.Date(), ".csv")),
              row.names = FALSE) #supplementary file 1.3
```

# Building an expression set object
Used for Limma analysis
```{r buildingEset}
# Constructing assay data
raw_data <- cdata %>%
  dplyr::select(starts_with("reporter_intensity")) %>%
  as.data.frame()
rownames(raw_data) <- cdata$Gene
names(raw_data) <- gsub("reporter_intensity_", "", names(raw_data))

# Constructing metadata
conditions_i <- data.frame(ID = names(raw_data))
conditions_i <- left_join(conditions_i, conditions)

# Constructing fdata
fdata <- cdata %>%
  dplyr::select(-starts_with("reporter_intensity")) %>%
  as.data.frame()

# Defining ID columns
rownames(fdata) <- fdata$Gene
rownames(conditions_i) <- conditions_i$ID
colnames(raw_data) <- conditions_i$ID

# Log2-transformation of raw_reporter_intensity values
raw_data_m <- log2(as.matrix(raw_data))
raw_data_m[is.infinite((raw_data_m))] <- NA
raw_data_m[is.na((raw_data_m))] <- NA

# Creating an expression set
raw_dataE <- ExpressionSet(assayData = raw_data_m,  #feature intensities (raw ion intensity for each channel)
                           phenoData = AnnotatedDataFrame(conditions_i), #metadata describing samples
                           featureData = AnnotatedDataFrame(fdata)) #metadata describing features (proteinID, etc)
validObject(raw_dataE)
rm(raw_data, raw_data_m, conditions_i, fdata)
```

# Batch effect removal

```{r removingBatchEffects}
batchcl_dataE <- raw_dataE
# CTX/INT samples corrected seaprately.
for (s in unique(conditions$sample))
{ batchcl_dataE_i <-  # Subset the data for the current sample and exclude missing values.
    batchcl_dataE[fData(batchcl_dataE)[, paste0("keep_", s)], 
                      batchcl_dataE$sample == s]
  
  exprs_available <- exprs(batchcl_dataE_i) # Extract the available (non-missing) expression values.
  
  non_missing_indices <- complete.cases(exprs_available) # Identify rows where all values are non-missing.
  
  exprs(batchcl_dataE_i)[non_missing_indices, ] <-  # Apply batch effect removal only to the non-missing values.
    removeBatchEffect(exprs_available[non_missing_indices, ], 
      batch = as.character(pData(batchcl_dataE_i)$rep), 
      design = model.matrix(~0 + as.character(pData(batchcl_dataE_i)$condition)))

  for (cn in colnames(batchcl_dataE_i))  # Write the batch-effect cleaned values back into the main data.
  {exprs(batchcl_dataE)[fData(batchcl_dataE)[, paste0("keep_", s)], cn] <- 
      exprs(batchcl_dataE_i)[, cn]}
  rm(cn)
}
rm(s)
```

# Data normalization
The vsn package from Wolfgang Huber is used to apply a variance stabilization normalization method on the log2 raw data.
```{r DataNormalization}
norm_dataE <- batchcl_dataE
for (s in unique(conditions$sample))
{ print(paste0(s))
  norm_input.data_i <- 2^exprs(norm_dataE[
    fData(norm_dataE)[, paste0("keep_", s)], 
    norm_dataE$sample == s])
  norm_dataE_i <- normalizeVSN(norm_input.data_i)
  for (cn in colnames(norm_dataE_i))
  {exprs(norm_dataE)[fData(norm_dataE)[, paste0("keep_", s)], cn] <- 
      norm_dataE_i[, cn]}
  rm(cn, norm_dataE_i)
  sdplot.object <- meanSdPlot(vsn2(norm_input.data_i), plot = FALSE)
  print(sdplot.object$gg + ggtitle(s))
  rm(sdplot.object, norm_input.data_i)
}
rm(s)
```

# PCA - principal component analysis
```{r CreatePCAplots, fig.height = 5}
sets <- list("Raw data" = raw_dataE, "Batch corrected data" = batchcl_dataE, "Normalised data" = norm_dataE)
PCA_data <- NULL
for (i in seq_along(sets)) {
  set <- sets[[i]]
  set.name <- names(sets)[i]
  for (s in unique(conditions$sample)) {
    set_s <- set[, set$sample == s]
    PCA_m <- t(na.omit(exprs(set_s)))
    if (length(PCA_m) > 50) {
      PCA <- prcomp(PCA_m, scale = FALSE)
      perc_var <-
        round(100 * PCA$sdev ^ 2 / #expressed as a percentage of the total dataset variance
                sum(PCA$sdev ^ 2), 1)
      PCA_data_i <-
        data.frame(PC1 = PCA$x[, 1],
                   PC2 = PCA$x[, 2],
                   PC3 = PCA$x[, 3],
                   PC1.var = perc_var[1],
                   PC2.var = perc_var[2],
                   sample = pData(set_s)$sample,
                   condition = pData(set_s)$condition,
                   rep = pData(set_s)$rep,
                   measurement = set.name)
      PCA_data <- bind_rows(PCA_data, PCA_data_i)
      rm(PCA_data_i)
    }
  }
  rm(s)
}
rm(i)
PCA_data$measurement <- factor(PCA_data$measurement, ordered = TRUE, levels = c("Raw data", "Batch corrected data", "Normalised data"))

PCA_data <- PCA_data %>%
  mutate(fraction = sub("_.*", "", condition),               # Create new columns for outline and fill
    genotype = if_else(str_ends(condition, "C"), "WT", "PT"))

unique_samples <- unique(PCA_data$sample)

#CTX and INT samples plotted separately
for (sample_name in unique_samples) {
  data_sub <- PCA_data %>% filter(sample == sample_name)
  
  p <- ggplot(data_sub, aes(PC1, PC2)) +
    geom_point(aes(fill = condition, colour = condition), shape = 21, size = 3, stroke = 1.2) +
    scale_fill_manual(values = c(
      "cytosol" = "#FFFFFF", 
      "cytosol_C" = "#D55E00",
      "membrane" = "#FFFFFF",
      "membrane_C" = "#0072B2")) +
    scale_colour_manual(values = c(
      "cytosol" = "#D55E00", 
      "cytosol_C" = "#D55E00",
      "membrane" = "#0072B2",
      "membrane_C" = "#0072B2")) +
    theme_minimal()+
    ggtitle(paste0("Principal Component analysis")) +
    xlab("PC1") +
    ylab("PC2") +
    facet_wrap(~ measurement + paste("PC1:", PC1.var, "% var - PC2:", PC2.var, "% var"), scales = "free", ncol = 4)
  
  ggsave(filename = file.path(dir_save, paste0("PCA_analysis_", script.version, "_", sample_name, "_", Sys.Date(), ".png")),
    plot = p, width = 11, height = 4, dpi = 600)
  
  write.csv(data_sub, file = file.path(dir_save, paste0("PCA_analysis_data_", script.version, "_", sample_name, "_", Sys.Date(), ".csv")), row.names = FALSE)
}

#both CTX and INT PCA in one plot 
p <- ggplot(PCA_data, aes(PC1, PC2)) +
  geom_point(aes(fill = condition, colour = condition), shape = 21, size = 3, stroke = 1.2) +
  scale_fill_manual(values = c(
    "cytosol" = "#FFFFFF", 
    "cytosol_C" = "#D55E00",
    "membrane" = "#FFFFFF",
    "membrane_C" = "#0072B2")) +
  scale_colour_manual(values = c(
    "cytosol" = "#D55E00", 
    "cytosol_C" = "#D55E00",
    "membrane" = "#0072B2",
    "membrane_C" = "#0072B2")) +
  theme_minimal()+
  ggtitle(paste0("Principal Component analysis")) +
  xlab("PC1") +
  ylab("PC2") +
  facet_wrap(sample~ measurement + paste("PC1:", PC1.var, "% var - PC2:", PC2.var, "% var"), scales = "free", ncol = 3)

ggsave(filename = file.path(dir_save, paste0("PCA_analysis_", script.version, "_",  Sys.Date(), ".png")),
  plot = p, width = 11, height = 6, dpi = 600)

write.csv(PCA_data, file = file.path(dir_save, paste0("PCA_analysis_data_", script.version, "_", Sys.Date(), ".csv")), row.names = FALSE)
```

# Merge modified data back into 'cdata'

```{r CreateCastedData}
cdata_i <- as.data.frame(2^exprs(batchcl_dataE))
names(cdata_i) <- paste0("batchcl_reporter_intensity_", names(cdata_i))
cdata_i$Gene <- rownames(cdata_i)
cdata <- left_join(cdata, cdata_i)
rm(cdata_i)

cdata_i <- as.data.frame(2^exprs(norm_dataE))
names(cdata_i) <- paste0("norm_reporter_intensity_", names(cdata_i))
cdata_i$Gene <- rownames(cdata_i)
cdata <- left_join(cdata, cdata_i)
rm(cdata_i)

write.csv(cdata, file = file.path(dir_save, paste0("Full_dataset_", script.version, ".csv")), row.names = FALSE) #supp. file 1.4
```

# Create tidy data
```{r CreateTidyData}
mdata <- NULL

mdata_i <- tidy(raw_dataE)
mdata_i <- mdata_i %>%
  mutate(value = 2 ^ value)
names(mdata_i)[1] <- "Gene"
names(mdata_i)[2] <- "ID"
mdata_i <- left_join(mdata_i, conditions)
mdata_i$measurement <- "raw_reporter_intensity"
mdata <- bind_rows(mdata, mdata_i)
rm(mdata_i)

mdata_i <- tidy(batchcl_dataE)
mdata_i <- mdata_i %>%
  mutate(value = 2 ^ value)
names(mdata_i)[1] <- "Gene"
names(mdata_i)[2] <- "ID"
mdata_i <- left_join(mdata_i, conditions)
mdata_i$measurement <- "batchcl_reporter_intensity"
mdata <- bind_rows(mdata, mdata_i)
rm(mdata_i)

mdata_i <- tidy(norm_dataE)
mdata_i <- mdata_i %>%
  mutate(value = 2 ^ value)
names(mdata_i)[1] <- "Gene"
names(mdata_i)[2] <- "ID"
mdata_i <- left_join(mdata_i, conditions)
mdata_i$measurement <- "norm_reporter_intensity"
mdata <- bind_rows(mdata, mdata_i)
rm(mdata_i)

mdata$condition <- factor(mdata$condition, ordered = TRUE, levels = c("cytosol_C", "cytosol", "membrane_C", "membrane"))
mdata$measurement <- factor(mdata$measurement, ordered = TRUE, levels = c("raw_reporter_intensity", "batchcl_reporter_intensity", "norm_reporter_intensity"))
```

# Data transformation overview
box plot of raw/batch corrected/normalised data across all 3 stages
```{r DataTransformationOverview}
customPlot2 <- list(
  theme_bw(base_size = 12), 
  scale_shape_manual(values = 21), 
  scale_fill_manual(values = c("#FFFFFF", "#D55E00", "#FFFFFF", "#0072B2")), 
  scale_colour_manual(values = c("#D55E00", "#D55E00", "#0072B2", "#0072B2")))

mdata$condition <- factor(mdata$condition, ordered = TRUE, levels = c("cytosol", "cytosol_C", "membrane", "membrane_C"))
new_labels <- c(raw_reporter_intensity = "Raw data",
  batchcl_reporter_intensity = "Batch corrected data",
  norm_reporter_intensity = "Normalised data")

# Individual plots by sample
samples <- unique(mdata$sample)
for (sample_name in samples) {
  data_no_ratio <- mdata %>% filter(sample == sample_name)
  p1 <- ggplot(data_no_ratio, aes(x=condition, y=log2(value), group = interaction(condition, rep))) +
    geom_boxplot(aes(fill = condition, colour=condition)) +
  stat_summary(
    data = subset(data_no_ratio, condition %in% c("cytosol_C", "membrane_C")),
    fun = median, geom = "crossbar",
    width = 0.7, colour = "white",
    fatten = 2, position = position_dodge(width = 0.75),
    aes(group = interaction(condition, rep)))+
    customPlot2 +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle(paste("Data Transformation Overview")) +
    theme_minimal()+
    facet_grid(. ~ measurement, labeller = labeller(measurement = new_labels)) +
  ylab("Log2(Reporter intensity)")
  try({
    ggsave(file.path(dir_save, paste0("Normalization_overview_",  sample_name, "_", script.version, "_", Sys.Date(), ".png")),
           plot = p1, width = 11, height = 4, dpi = 150)
  }, silent = TRUE)
  while (!is.null(dev.list())) dev.off()
  
  try({
    write.csv(data_no_ratio, file.path(dir_save, paste0("Normalization_overview_data_",  sample_name, "_", script.version, "_", Sys.Date(), ".csv")), row.names = FALSE)
  }, silent = TRUE)
  
  rm(p1)
  gc()
}

# Plot both samples together
data_no_ratio <- data_no_ratio %>%
  pivot_wider(names_from = measurement,
    values_from = value)
data_no_ratio <- na.omit(data_no_ratio)

write.csv(data_no_ratio, file.path(dir_save, paste0("Normalization_overview_data_cleaned_",  sample_name, "_", script.version, "_", Sys.Date(), ".csv")), row.names = FALSE)

data_no_ratio <- mdata %>% filter(!grepl("ratio", measurement))
  p3 <- ggplot(data_no_ratio, aes(x=condition, y=log2(value), group = interaction(condition, rep))) +
    geom_boxplot(aes(fill = condition, colour=condition)) +
  stat_summary(
    data = subset(data_no_ratio, condition %in% c("cytosol_C", "membrane_C")),
    fun = median,
    geom = "crossbar",
    width = 0.7,
    colour = "white",
    fatten = 2,
    position = position_dodge(width = 0.75),
    aes(group = interaction(condition, rep)))+
    customPlot2 +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    ggtitle(paste("Data Transformation Overview")) +
    theme_minimal()+
    facet_grid(sample ~ measurement, labeller = labeller(measurement = new_labels)) +
  ylab("Log2(Reporter intensity)")
  try({
    ggsave(file.path(dir_save, paste0("Normalization_overview_",  script.version, "_", Sys.Date(), ".png")),
           plot = p3, width = 11, height = 4, dpi = 600)
  }, silent = TRUE)
  while (!is.null(dev.list())) dev.off()
  
  try({
    write.csv(data_no_ratio, file.path(dir_save, paste0("Normalization_overview_data_",  script.version, "_", Sys.Date(), ".csv")),
              row.names = FALSE)
  }, silent = TRUE)
```

# CV overview (coefficient of variation)
to check if data normalisation has stabilised variance independent of mean
```{r CVOverview, fig.height = 6}
cv.data <- mdata %>%
  group_by(Gene, measurement, sample, condition) %>%
  summarise(CV = sd(value, na.rm = TRUE) / mean(value, na.rm = TRUE))
cv.data.all <- mdata %>%
  mutate(condition = "all") %>%
  group_by(Gene, measurement, sample, condition) %>%
  summarise(CV = sd(value, na.rm = TRUE) / mean(value, na.rm = TRUE))
cv.data <- bind_rows(cv.data, cv.data.all)
rm(cv.data.all)
ggplot(data = cv.data, aes(condition, CV * 100)) +
  geom_violin() +
  geom_boxplot(width = 0.1) +

  ggtitle("CV overview") +
  ylab("CV [%]") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  facet_grid(sample ~ measurement)
ggsave(file.path(dir_save, paste0("CV_overview_", script.version, ".png")), width = 12, height = 10)
```

# LIMMA analysis
A protein is considered a 'hit', if the false discovery rate is smaller 0.05 and an absolute log2 fold-change of at least log2(1.5) is observed. 
If separate datasets (sample grouping) are being analysed in the same dataset, Limma used to compare within each dataset, not as a whole
```{r LimmaInitialization}
limma_results <- NULL
```

## Run limma analysis
CTX and INT samples run separately

```{r LimmaRun}
for (s in unique(conditions$sample))
{ limma_data <- norm_dataE[
    fData(norm_dataE)[, paste0("keep_", s)], 
    norm_dataE$sample == s]
  limma_weights <- 
    exprs(raw_dataE[fData(raw_dataE)[, paste0("keep_", s)], 
                    raw_dataE$sample == s])
  #compare effect of genotype and fraction separation
  comparison.list <- c("cytosol - cytosol_C", "membrane - membrane_C", "membrane - cytosol", "membrane_C - cytosol_C")
  limma_weights <- ifelse(is.na(limma_weights), 0.01, 1) #assign low weight to missing values
  limma.cond <- factor(pData(limma_data)$condition, ordered = FALSE)
  limma.rep <- factor(pData(limma_data)$rep, ordered = FALSE)
  design.matrix <- model.matrix( ~  0 + limma.cond + limma.rep)
  colnames(design.matrix) <- gsub("limma.cond", "", colnames(design.matrix))
  
  limma.object <- eBayes(
    contrasts.fit(lmFit(limma_data, design = design.matrix, weights = limma_weights), 
      makeContrasts(contrasts = comparison.list, levels = design.matrix))
  )
  for (comparison in comparison.list){
    limma_results_i <- limma::topTable(limma.object, 
                           sort.by = "t",  
                           coef = comparison, 
                           number = Inf)
    names(limma_results_i)[grep("P.Value", names(limma_results_i))] <- "pvalue.limma"
    names(limma_results_i)[grep("adj.P.Val", names(limma_results_i))] <- "fdr.limma"
    limma_results_i <- subset(limma_results_i, !is.na(logFC))
    fdr_res <- NULL#also use FDRtool to calculate false discovery rate to independently check Limma results
    try(fdr_res <- fdrtool(limma_results_i$t, plot = FALSE, verbose = FALSE))
    if (!is.null(fdr_res))
    {limma_results_i$pvalue.fdrtool <- fdr_res$pval
      limma_results_i$qval.fdrtool <- fdr_res$qval
      limma_results_i$lfdr.fdrtool <- fdr_res$lfdr
    }
    else {
      limma_results_i$pvalue.fdrtool <- 1
      limma_results_i$qval.fdrtool <- 1
      limma_results_i$lfdr.fdrtool <- 1}
    
    comparison.label <- gsub("\\) - \\(", ") against (", comparison)
    comparison.label <- gsub(" - ", " vs ", comparison.label)
    limma_results_i$comparison <- comparison
    limma_results_i$comparison.label <- comparison.label
    limma_results_i$sample <- s
    limma_results <- bind_rows(limma_results, limma_results_i)
    rm(limma_results_i, fdr_res)
  }
  rm(comparison.list, comparison, design.matrix, limma.cond, limma.rep, limma_data, limma_weights, limma.object)
}
rm(s)
```

## Fdr statistics comparison limma vs fdrtool
```{r LimmaFDRcomparison, fig.height = 7}
customPlot3 <- list(
  theme_bw(base_size = 12), 
  scale_shape_manual(values = 21), 
  scale_fill_manual(values = c("#e41a1c", "#009E73")), 
  scale_colour_manual(values = c("#e41a1c", "#009E73")))

samples <- unique(limma_results$sample)
for (s in samples) {
  # Subset data for current sample
  dat_sub <- subset(limma_results, sample == s)
  
  # Plot 1: t vs fdr
  p1 <- ggplot(dat_sub, aes(abs(t))) + 
    geom_line(aes(y = fdr.limma, colour = "limma - adj.P.Val")) + 
    geom_line(aes(y = qval.fdrtool, colour = "fdrtool - qval")) + 
    geom_point(aes(y = fdr.limma, colour = "limma - adj.P.Val")) + 
    geom_point(aes(y = qval.fdrtool, colour = "fdrtool - qval")) + 
    facet_wrap(~ paste0(sample, ": ", comparison.label), scales = "free_x") + 
    ylab("fdr") + 
    customPlot3 +
    ggtitle(paste0("FDR comparison limma vs fdrtool - Sample: ", s))
  
  ggsave(filename = file.path(dir_save, paste0("t_vs_fdr_limma_vs_fdrtool_", s, "_", script.version, "_", Sys.Date(), ".png")),
    plot = p1, width = 20, height = 13.5, dpi = 300)
  
  # Plot 2: p-value histogram
  p2 <- ggplot(dat_sub) + 
    geom_histogram(aes(pvalue.limma, alpha = 0.5, fill = "limma"), bins = 40) + 
    geom_histogram(aes(pvalue.fdrtool, alpha = 0.5, fill = "fdrtool"), bins = 40) + 
    guides(alpha = FALSE) + 
    xlab("p-value") + 
    facet_wrap(~ paste0(sample, ": ", comparison.label), scales = "free_y") +
    coord_cartesian(xlim = c(0, 1)) +
    customPlot3 +
    ggtitle(paste0("p-value histogram limma vs fdrtool - Sample: ", s))
  
  ggsave(filename = file.path(dir_save, paste0("p-value_histogram_limma_vs_fdrtool_", s, "_", script.version, "_", Sys.Date(), ".png")),
    plot = p2, width = 20, height = 13.5, dpi = 600)
}
```

## Hit annotation - All proteins
```{r LimmaHitAnnotation}
limma_results$hit_annotation_method <- NA
fdr_hit_threshold <- 0.05 #FDR threshold
fc_hit_threshold <- 1.5 #Fold-change threshold
limma_results$pvalue <- NA
limma_results$fdr <- NA
limma_results$comparison.old <- limma_results$comparison
limma_results$comparison <- with(limma_results, paste0(sample, ": ", comparison))
for (comparison in unique(limma_results$comparison))
{limma_hits <- 
    nrow(limma_results[limma_results$comparison == comparison & 
                         limma_results$fdr.limma <= fdr_hit_threshold, ])
  fdrtool_hits <- 
    nrow(limma_results[limma_results$comparison == comparison & 
                         limma_results$qval.fdrtool <= fdr_hit_threshold, ])
  if (limma_hits >= fdrtool_hits) {
    limma_results$hit_annotation_method[limma_results$comparison == comparison] <- "limma"}
  if (fdrtool_hits > limma_hits)
  {limma_results$hit_annotation_method[limma_results$comparison == comparison] <- "fdrtool"}
  rm(limma_hits, fdrtool_hits)
}

table(limma_results$hit_annotation_method)
limma_results$hit_annotation_method[!grepl("[(,)]", limma_results$comparison)] <- "limma"

limma_results$pvalue[limma_results$hit_annotation_method == "limma"] <- 
  limma_results$pvalue.limma[limma_results$hit_annotation_method == "limma"]
limma_results$fdr[limma_results$hit_annotation_method == "limma"] <- 
  limma_results$fdr.limma[limma_results$hit_annotation_method == "limma"]

limma_results$pvalue[limma_results$hit_annotation_method == "fdrtool"] <- 
  limma_results$pvalue.fdrtool[limma_results$hit_annotation_method == "fdrtool"]
limma_results$fdr[limma_results$hit_annotation_method == "fdrtool"] <- 
  limma_results$qval.fdrtool[limma_results$hit_annotation_method == "fdrtool"]

limma_results$hit <- 
  with(limma_results, ifelse(fdr <= fdr_hit_threshold & abs(logFC) >= 
                               log2(fc_hit_threshold), TRUE, FALSE))
limma_results$hit_annotation <- with(limma_results,
  ifelse(fdr <= fdr_hit_threshold & abs(logFC) >= log2(fc_hit_threshold),
         "hit", "no hit"))
limma_results$hit_annotation <- 
  factor(limma_results$hit_annotation, 
         ordered = TRUE, 
         levels = c("hit", "no hit"))
caption.text_DE = glue("hit: fdr <= {fdr_hit_threshold} & absolute log2(fold-change) >= log2({fc_hit_threshold})")

with(limma_results, table(sample, comparison, hit_annotation))
limma_results$comparison <- limma_results$comparison.old
limma_results$comparison.old <- NULL

# Save limma results : supp file 1.5
write.csv(limma_results, file.path(dir_save, paste0("Limma_results_ALL_", script.version,"_", Sys.Date(), ".csv")), row.names = FALSE) 
```

#human protein atlas - neuron elevated
```{r HumanProteinAtlasBrainExpression}
#filter hits to keep any protein listed in the Human Protein Atlast as elevated expression in neurons
## this dataset is 3795 proteins published here https://www.proteinatlas.org/humanproteome/brain 

ProteinAtlasWhole <- read.xlsx("C:/Users/laura/OneDrive - Cardiff University/Documents/3 - PhD/Proteomics/human protein atlas dataset all neuron total elevated.xlsx", sheet = 1)

limma_wholebrain_removed<-limma_results%>% #save dataset of hits being removed (non-neuronal enriched)
  dplyr::filter(!(Gene %in% ProteinAtlasWhole$Gene)&
         !(Protein.ID %in% ProteinAtlasWhole$Uniprot) )%>%
  dplyr::filter(comparison %in% c("cytosol - cytosol_C", "membrane - membrane_C"))%>% #select Patient-Control comparisons
  dplyr::filter(hit == "TRUE")
write.csv(limma_wholebrain_removed, file.path(dir_save, paste0("Limma_results_Neuron removed_", script.version,"_", Sys.Date(), ".csv")), row.names = FALSE) 

limma_results<-limma_results%>% #filter Limma results for proteins in Human Protein Atlas
  dplyr::filter(Gene %in% ProteinAtlasWhole$Gene|
         Protein.ID %in% ProteinAtlasWhole$Uniprot )

write.csv(limma_results, file.path(dir_save, paste0("Limma_results_Neuron enriched_", script.version,"_", Sys.Date(), ".csv")), row.names = FALSE) #supp file 1.6

unique_hits<-as.data.frame(limma_results)%>% #to see unique list of differentially expressed proteins
  filter(hit==TRUE)%>%
  dplyr::select(Gene, sample, comparison, logFC)%>%
  distinct()%>%
  mutate(direction = case_when(
    logFC > 0 ~ "Up",
    logFC < 0 ~ "Down",
    TRUE ~ "NoChange"   # optional: handles cases where logFC == 0 or NA
  ))
sample_only <- unique_hits %>%
  group_by(sample) %>%
  filter(comparison %in% c("membrane - cytosol","membrane_C - cytosol_C"))%>%
  distinct()%>%
  summarise(count = n(), .groups = "drop")
fraction_only <- unique_hits %>%
  group_by(sample, comparison) %>%
  distinct()%>%
  summarise(count = n(), .groups = "drop")
fraction_direction <- unique_hits %>%
  group_by(sample, comparison, direction) %>%
  distinct()%>%
  summarise(count = n(), .groups = "drop")

print(fraction_only)
print(fraction_direction)
```

# Volcano-plot
```{r LimmaVolcanoPlot, fig.height = 7}
samples <- unique(limma_results$sample)
customPlot_DE2 <- list(
  theme_bw(base_size = 12), 
  scale_shape_manual(values = c(16, 17, 15, 3, 7, 8)), 
  scale_colour_manual(values = c("hit" = "#1f78b4", "no hit" = "#555555")),
  scale_alpha_manual(values = c("hit" = 1,  "no hit" = 0.1)))


#CTX/INT and volcano plots together - effect of genotype
dat_sample <- limma_results%>%
  filter(comparison %in% c("cytosol - cytosol_C", "membrane - membrane_C"))
  
volcano_labels <- dat_sample %>%
  group_by(sample, comparison.label) %>%
  summarise(up_hits = sum(hit_annotation == "hit" & logFC > 0),
          down_hits = sum(hit_annotation == "hit" & logFC < 0),
          total_points = n(), .groups = "drop")

p <- ggplot(dat_sample, aes(logFC, -log10(pvalue), colour = hit_annotation, alpha = hit_annotation)) +
  geom_vline(xintercept = 0, colour = "lightgrey") +
  geom_hline(yintercept = -Inf, colour = "black") +
  geom_vline(xintercept = -Inf, colour = "black") +
  geom_point() +
  geom_text(aes(label = gsub("[|].+", "", Gene)), colour = "black",
    show.legend = FALSE,
    data = subset(dat_sample, hit_annotation == "hit"),
    vjust = 0, nudge_y = 0.1, size = 2, check_overlap = TRUE) +
  facet_grid(sample~comparison.label, scales ="fixed") +
  xlab("log2(fold-change)") +
  customPlot_DE2 +
  theme_classic() +
  theme(strip.background = element_blank(),)+
  geom_text(data = volcano_labels, aes(x = -3.5, y = 0,
      label = paste0(
        "Upregulated hits: ", up_hits, "\n",
        "Downregulated hits: ", down_hits, "\n",
        "Total: ", total_points)),
    hjust = 0, vjust = 0, inherit.aes = FALSE, size = 2.5, colour = "black") +
  coord_cartesian(clip = "off")+
  labs(colour = "Hit annotation", alpha = "Hit annotation") 

ggsave(filename = file.path(dir_save, paste0("Volcano_plot_genotype_", script.version, "_", Sys.Date(), ".png" )),
  plot = p, width =10, height = 6)

#CTX/INT and volcano plots together - effect of fraction
dat_sample <- limma_results%>%
  filter(comparison %in% c("membrane - cytosol", "membrane_C - cytosol_C"))
  
volcano_labels <- dat_sample %>%
  group_by(sample, comparison.label) %>%
  summarise(up_hits = sum(hit_annotation == "hit" & logFC > 0),
          down_hits = sum(hit_annotation == "hit" & logFC < 0),
          total_points = n(), .groups = "drop")

p <- ggplot(dat_sample, aes(logFC, -log10(pvalue), colour = hit_annotation, alpha = hit_annotation)) +
  geom_vline(xintercept = 0, colour = "lightgrey") +
  geom_hline(yintercept = -Inf, colour = "black") +
  geom_vline(xintercept = -Inf, colour = "black") +
  geom_point() +
  geom_text(aes(label = gsub("[|].+", "", Gene)), colour = "black",
    show.legend = FALSE,
    data = subset(dat_sample, hit_annotation == "hit"),
    vjust = 0, nudge_y = 0.1, size = 2, check_overlap = TRUE) +
  facet_grid(sample~comparison.label, scales ="fixed") +
  xlab("log2(fold-change)") +
  customPlot_DE2 +
  theme_classic() +
  theme(strip.background = element_blank(),)+
  geom_text(data = volcano_labels, aes(x = -3.5, y = 0,
      label = paste0(
        "Upregulated hits: ", up_hits, "\n",
        "Downregulated hits: ", down_hits, "\n",
        "Total: ", total_points)),
    hjust = 0, vjust = 0, inherit.aes = FALSE, size = 2.5, colour = "black") +
  coord_cartesian(clip = "off")+
  labs(colour = "Hit annotation", alpha = "Hit annotation") 

ggsave(filename = file.path(dir_save, paste0("Volcano_plot_fraction_", script.version, "_", Sys.Date(), ".png" )),
  plot = p, width =10, height = 4, dpi=600)

#effect of genotype in membrane
dat_sample <- limma_results%>%
  filter(comparison %in% c("membrane - membrane_C"))
  
volcano_labels <- dat_sample %>%
  group_by(sample, comparison.label) %>%
  summarise(up_hits = sum(hit_annotation == "hit" & logFC > 0),
          down_hits = sum(hit_annotation == "hit" & logFC < 0),
          total_points = n(), .groups = "drop")

p <- ggplot(dat_sample, aes(logFC, -log10(pvalue), colour = hit_annotation, alpha = hit_annotation)) +
  geom_vline(xintercept = 0, colour = "lightgrey") +
  geom_hline(yintercept = -Inf, colour = "black") +
  geom_vline(xintercept = -Inf, colour = "black") +
  geom_point() +
  geom_text(aes(label = gsub("[|].+", "", Gene)), colour = "black",
    show.legend = FALSE,
    data = subset(dat_sample, hit_annotation == "hit"),
    vjust = 0, nudge_y = 0.1, size = 2, check_overlap = TRUE) +
  facet_grid(sample~comparison.label, scales ="fixed") +
  xlab("log2(fold-change)") +
  customPlot_DE2 +
  theme_classic() +
  theme(strip.background = element_blank(),)+
  geom_text(data = volcano_labels, aes(x = -3.2, y = 0,
      label = paste0(
        "Upregulated hits: ", up_hits, "\n",
        "Downregulated hits: ", down_hits, "\n",
        "Total: ", total_points)),
    hjust = 0, vjust = 0, inherit.aes = FALSE, size = 2.5, colour = "black") +
  coord_cartesian(clip = "off")+
  labs(colour = "Hit annotation", alpha = "Hit annotation") 

ggsave(filename = file.path(dir_save, paste0("Volcano_plot_genotype_membrane_", script.version, "_", Sys.Date(), ".png" )),
  plot = p, width =8, height = 4, dpi=600)

#effect of genotype in cytosol
dat_sample <- limma_results%>%
  filter(comparison %in% c("cytosol - cytosol_C"))
  
volcano_labels <- dat_sample %>%
  group_by(sample, comparison.label) %>%
  summarise(up_hits = sum(hit_annotation == "hit" & logFC > 0),
          down_hits = sum(hit_annotation == "hit" & logFC < 0),
          total_points = n(), .groups = "drop")

p <- ggplot(dat_sample, aes(logFC, -log10(pvalue), colour = hit_annotation, alpha = hit_annotation)) +
  geom_vline(xintercept = 0, colour = "lightgrey") +
  geom_hline(yintercept = -Inf, colour = "black") +
  geom_vline(xintercept = -Inf, colour = "black") +
  geom_point() +
  geom_text(aes(label = gsub("[|].+", "", Gene)), colour = "black",
    show.legend = FALSE,
    data = subset(dat_sample, hit_annotation == "hit"),
    vjust = 0, nudge_y = 0.1, size = 2, check_overlap = TRUE) +
  facet_grid(sample~comparison.label, scales ="fixed") +
  xlab("log2(fold-change)") +
  customPlot_DE2 +
  theme_classic() +
  theme(strip.background = element_blank(),)+
  geom_text(data = volcano_labels, aes(x = -2.7, y = 0,
      label = paste0(
        "Upregulated hits: ", up_hits, "\n",
        "Downregulated hits: ", down_hits, "\n",
        "Total: ", total_points)),
    hjust = 0, vjust = 0, inherit.aes = FALSE, size = 2.5, colour = "black") +
  coord_cartesian(clip = "off")+
  labs(colour = "Hit annotation", alpha = "Hit annotation") 

ggsave(filename = file.path(dir_save, paste0("Volcano_plot_genotype_cytosol_", script.version, "_", Sys.Date(), ".png" )),
  plot = p, width =8, height = 5, dpi=600)
```

## Fold-change correlation between INT and CTX
### all hits
```{r LimmaFoldChangeCorrelationSamples, fig.height = 7}
customPlot_DE_cor2 <- list(
  theme_bw(base_size = 12), 
  labs(caption = caption.text_DE), 
  scale_colour_manual(values = c("hit INT - hit CTX" = "#CC79A7", "hit INT - no hit CTX" = "#0072B2", "no hit INT - hit CTX" = "#D55E00", "no hit INT - no hit CTX" = "#555555")),
  scale_alpha_manual(values = c("hit INT - hit CTX" = 1,  "hit INT - no hit CTX" = 1, "no hit INT - hit CTX" = 1,  "no hit INT - no hit CTX" = 0.1)))

sample.cor.data <- limma_results %>%
  dplyr::filter(sample %in% c("INT", "CTX")) %>%
  dplyr::filter(comparison %in% c("cytosol - cytosol_C", "membrane - membrane_C")) %>%
  group_by(Gene, comparison) %>%
  dplyr::select(Gene, comparison, comparison.label, logFC, sample) %>%
  spread(key = sample, value = logFC)
sample.cor.data_i <- limma_results %>%
  group_by(Gene, comparison) %>%
  dplyr::select(Gene, comparison, comparison.label, hit_annotation, sample) %>%
  mutate(hit_annotation = gsub("enriched ", "", hit_annotation)) %>%
  spread(key = sample, value = hit_annotation)
names(sample.cor.data_i)[c(4,5)] <- paste0("hit_", names(sample.cor.data_i)[c(4,5)])
sample.cor.data <- left_join(sample.cor.data, sample.cor.data_i)
rm(sample.cor.data_i)
sample.cor.data <- na.omit(sample.cor.data)
sample.cor.data$hit <- paste(sample.cor.data$hit_INT, "INT -", sample.cor.data$hit_CTX, "CTX")
sample.cor.data$hit <- factor(sample.cor.data$hit, ordered = TRUE,
                       levels = c("hit INT - hit CTX",
                                  "hit INT - no hit CTX",
                                  "no hit INT - hit CTX",
                                  "no hit INT - no hit CTX"))
no.of.comparisons <- length(unique(sample.cor.data$comparison))
max.comp.nchar <- max(nchar(unique(sample.cor.data$comparison)))
max.comp.nchar <- ifelse(max.comp.nchar <= 70, 70, max.comp.nchar)
gr.width.multiplikator <- round(max.comp.nchar / 70 * 5, digits = 2)
gr.width <- 4 + ceiling(sqrt(no.of.comparisons)) * gr.width.multiplikator
gr.height <- 4 + floor(sqrt(no.of.comparisons)) * 5
ggplot(data = sample.cor.data, aes(INT, CTX, colour = hit, alpha = hit)) +
  geom_abline() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point() +
  facet_wrap(~ comparison.label, scale = "free") +
  geom_text(aes(label = gsub("[|].+", "", Gene)), colour = "black", show.legend = FALSE, 
            data = subset(sample.cor.data, hit != "no hit INT - no hit CTX"),
            vjust = 0, nudge_y = 0.1, size = 2, check_overlap = TRUE) +
  customPlot_DE_cor2 +
  ggtitle("Fold-change correlation plot INT vs CTX")
ggsave(file.path(dir_save, paste0("Fold_change_correlation_INT_vs_CTX_", script.version, ".png")), width = gr.width, height = gr.height)
write.csv(sample.cor.data, file = file.path(dir_save, paste0("Fold_change_correlation_data_INT_vs_CTX_", script.version, ".csv")), row.names = FALSE)
```

```{r LimmaFoldChangeCorrelationSamples}
library(ggrepel)
## Save limma results
sample.cor.data <- sample.cor.data %>%
  mutate(
    direction = case_when(
      hit == "hit INT - hit CTX" & INT > 0 & CTX > 0 ~ "Up in both",
      hit == "hit INT - hit CTX" & INT < 0 & CTX < 0 ~ "Down in both",
      hit == "hit INT - hit CTX" & INT > 0 & CTX < 0 ~ "Up INT / Down CTX",
      hit == "hit INT - hit CTX" & INT < 0 & CTX > 0 ~ "Down INT / Up CTX",
      TRUE ~ "Other"))

direction_cols <- c("Up in both"= "#7EEC57",
  "Down in both"         = "#F8766D",
  "Up INT / Down CTX"    = "#00BFC4",
  "Down INT / Up CTX"    = "#8494FF",
  "Other"                = "#555555")
direction_alpha <- c("Up in both" = 1,
  "Down in both"         = 1,
  "Up INT / Down CTX"    = 1,
  "Down INT / Up CTX"    = 1,
  "Other"                = 0.1)
legend_groups <- c("Up in both",
  "Down in both",
  "Up INT / Down CTX",
  "Down INT / Up CTX")
customPlot_DE_cor2 <- list(
  theme_minimal(),
  labs(x = "INT log2(fold-change)",
    y = "CTX log2(fold-change)"),
  scale_colour_manual(values = direction_cols, breaks = legend_groups),
  scale_alpha_manual(values = direction_alpha, breaks = legend_groups))

ggplot(data = sample.cor.data,
       aes(INT, CTX, colour = direction, alpha = direction)) +
  geom_abline(colour = "grey50", linetype = "dotted") +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_point() +
  facet_wrap(~ comparison.label, scale = "free", ncol=1) +
  geom_text_repel(
    aes(label = gsub("[|].+", "", Gene)),
    colour = "black",
    data = subset(sample.cor.data, hit == "hit INT - hit CTX"),
    size = 2.5, max.overlaps = Inf, force = 1.5,
    box.padding = 0.3, point.padding = 0.1) +
  customPlot_DE_cor2 +
  guides(alpha = "none",
    colour = guide_legend(override.aes = list(size = 4),nrow = 1, byrow = TRUE))+   # Bigger colour legend dots
  coord_cartesian(xlim = c(-3, 3.5), ylim = c(-2, 3) )+
  labs(x = "Inhibitory log2(fold-change)",
       y = "Excitatory log2(fold-change)") +
  theme(legend.position = "bottom")

ggsave(file.path(dir_save, paste0("Fold_change_correlation_INT_vs_CTX_hits only v2_", script.version, ".png")), width = 6.5, height = 10, dpi =600)
```

```{r LimmaSave}
write.csv(limma_results, file.path(dir_save, paste0("Limma_results_filtered_", script.version,"_", Sys.Date(), ".csv")), row.names = FALSE)
```

# Save R workspace

```{r}
save.image(file.path(dir_save, paste0("Workspace_", script.version, ".RData")))
# load(file.path(dir_save, paste0("Workspace_", script.version, ".RData")))
```

# GO term enrichment analysis
## Preparation
```{r GO_IDLUT}
calculateFE <- function(GeneRatio, BgRatio) #function to calculate fold enirchment/Odds ratio
{ GeneRatio <- as.numeric(unlist(str_split(GeneRatio, pattern = "/"))) #ratio for each gene
  BgRatio <- as.numeric(unlist(str_split(BgRatio, pattern = "/"))) #ratio for background
  odds_ratio <- (GeneRatio[1] / GeneRatio[2]) / (BgRatio[1] / BgRatio[2]) #odds ratio
  return(odds_ratio)}
ID_LUT <- bitr(geneID = cdata$Gene, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
names(ID_LUT) <- c("Gene", "ENTREZID")
```

## Differential expression analysis
```{r GO_DE}
limma_results <- left_join(limma_results, ID_LUT) #join GO ID columns to Limma data
ego_data <- limma_results %>%
  dplyr::filter(!is.na(ENTREZID), hit_annotation %in% c("hit")) %>% #only use proteins with ID and "hits"
  dplyr::select(ENTREZID, sample, comparison, logFC) %>%
  distinct() %>% 
  mutate(comparison = gsub("[.]", "_", comparison)) %>%
  mutate(grouping = paste(sample, comparison, sep = "___")) %>% #create new grouping variable
  mutate(direction = ifelse(logFC > 0, "upregulated", "downregulated")) %>% #define up or downregulated
  dplyr::select(ENTREZID, grouping, direction)
```

### Cellular compartment
```{r GO_DE_CC}
ego_results_DE_CC <- NULL
try(ego_results_DE_CC <- compareCluster(
  ENTREZID ~ grouping + direction, #grouping by sample and up/downregulated
  data = ego_data,  #use dataset generated with ENTREZID
  fun = "enrichGO", #clusterProfiler::enrichGO()
  OrgDb = org.Hs.eg.db, #human Org database
  keyType = "ENTREZID",
  ont = "CC", #cellular compartment
  readable = TRUE,
  universe = ID_LUT$ENTREZID)) #background genes

if (!is.null(ego_results_DE_CC)) {
  ego_results_DE_CC_table <- ego_results_DE_CC@compareClusterResult %>% 
    mutate(comparison = gsub(".+___", "", grouping),
      sample = gsub("___.+", "", grouping)) %>%
    group_by(ID, sample, comparison, direction) %>%
    mutate(odds_ratio = calculateFE(GeneRatio, BgRatio)) %>%
    ungroup()
  
  ego_results_DE_CC_table$Description <- factor(
    ego_results_DE_CC_table$Description,
    levels = unique(rev(ego_results_DE_CC_table$Description)))

  ego_sub <- ego_results_DE_CC_table %>%
    mutate(is_cyt_mem = grepl("cytosol\\s*-\\s*cytosol_C", grouping, ignore.case = TRUE) |
                   grepl("membrane\\s*-\\s*membrane_C", grouping, ignore.case = TRUE)) %>%
    group_by(sample, comparison, direction) %>%
    arrange(p.adjust)%>% #arrange by smallest adjusted p-value first
    slice_head(n = 10) %>% # Keep top 10 per group
    group_by(is_cyt_mem) %>%
    arrange(p.adjust, .by_group = TRUE) %>%
    mutate(Description_reordered = factor(Description, levels = unique(Description))) %>%
    ungroup()
}

#horizontal bar plots 
df<-ego_results_DE_CC_table
df<-df%>%
  filter(comparison %in% c("membrane - membrane_C", "cytosol - cytosol_C"))

df$logP <- -log10(df$p.adjust)
df <- df %>%
  arrange(desc(logP), .by_group = TRUE)%>%
  group_by(Cluster) %>%
  slice_max(order_by = logP, n = 10, with_ties = FALSE)

#define custom plot colours
df$direction <- factor(df$direction, levels = c("downregulated", "upregulated"))
direction_colors <- c("downregulated" = "#F8766D", "upregulated" = "#00B358")
df <- df %>%
  mutate(direction_colors = case_when(
    str_detect(tolower(Cluster), "cytosol") & direction == "downregulated" ~ "Downregulated cytosol",
    str_detect(tolower(Cluster), "cytosol") & direction == "upregulated"   ~ "Upregulated cytosol",
    str_detect(tolower(Cluster), "membrane") & direction == "downregulated" ~ "Downregulated membrane",
    str_detect(tolower(Cluster), "membrane") & direction == "upregulated"   ~ "Upregulated membrane",
    TRUE ~ "other"))

ego_results_DE_CC_table2 <- ego_results_DE_CC_table %>%
  filter(comparison %in% c("cytosol - cytosol_C", "membrane - membrane_C")) %>%
  group_by(sample, comparison, direction) %>%
  arrange(p.adjust, .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  # Convert all factors to character before saving
  mutate(across(where(is.factor), as.character))

write.csv(ego_results_DE_CC_table2,
  file.path(dir_save, paste0("GO_enrichment_DE_CC_limma_results_top10_", script.version, ".csv")),
  row.names = FALSE)

#count how many times each gene appears in GO terms
gene_counts_CC <- ego_results_DE_CC_table %>%
  filter(comparison %in% c("cytosol - cytosol_C", "membrane - membrane_C"))%>%
  group_by(sample)%>%
  separate_rows(geneID, sep = "/") %>% # split by "/" 
  mutate(geneID = str_trim(geneID)) %>%
  count(geneID, sort = TRUE)%>% #count how many times each gene appears
  pivot_wider(
    names_from = sample,
    values_from = n,
    values_fill = 0 )

write.csv(gene_counts_CC, file.path(dir_save, paste0("GO_gene_counts_CC_", script.version, ".csv")), row.names = FALSE)

#define custom plot colours
df$direction <- factor(df$direction, levels = c("downregulated", "upregulated"))
direction_colors <- c("downregulated" = "#F8766D", "upregulated" = "#00B358")
df <- df %>%
  mutate(direction_colors = case_when(
    sample == "CTX" & direction == "downregulated" ~ "Downregulated CTX",
    sample == "CTX"  & direction == "upregulated"   ~ "Upregulated CTX",
    sample == "INT"  & direction == "downregulated" ~ "Downregulated INT",
    sample == "INT"  & direction == "upregulated"   ~ "Upregulated INT",
    TRUE ~ "other" 
  ))
outline_color <- c("Downregulated CTX"   = "#F8766D",     
                    "Upregulated CTX"     = "#00B358",     
                    "Downregulated INT"  = "#F8766D",     
                    "Upregulated INT"    = "#00B358" )
fill_color <- c("Downregulated CTX"   = "#F8766D",     
                "Upregulated CTX"     = "#00B358",       
                "Downregulated INT"  = "white",    
                "Upregulated INT"    = "white")
df$direction_colors <- factor(df$direction_colors, levels = c("Downregulated CTX", "Upregulated CTX", "Downregulated INT", "Upregulated INT"))

#plot all together - separated by cytosol/membrane
for (s in unique(df$comparison)) {
  plot_data <- df %>%      
  filter(comparison == s)
  
p <- ggplot(plot_data, aes(x = logP, y = reorder_within(Description, logP, Cluster), fill = direction_colors, color=direction_colors)) +
            geom_col(width = 0.4) +
            scale_fill_manual(values = fill_color) +
            scale_color_manual(values = outline_color) +
            scale_y_reordered() +
            facet_wrap(Cluster ~ ., scales = "free", space = "free_y", ncol=2) +
            theme_minimal() +
            labs(x = expression(-log[10](p.adjust)),
              y = NULL, title = "i) Cellular Compartment") +
            theme(panel.background = element_rect(fill = "white", color = NA), 
              plot.background = element_rect(fill = "white", color = NA),   
              axis.text.y = element_text(color = "black", size = 7, lineheight = 0.9),
              axis.text.x = element_text(color = "black"),
              plot.title = element_text(hjust = 0.5),
              strip.text.y = element_text(angle = 0, hjust = 0),
              panel.grid.major = element_blank(),  
              panel.grid.minor = element_blank(),
              strip.background = element_blank(),
              strip.text.x = element_blank(),
              legend.title = element_blank())+
          theme(legend.position = "bottom")
          print(p)
          ggsave(file.path(dir_save, paste0("GO_enrichment_DE_CC_barplot_", s, "-",  script.version, "_", Sys.Date(), ".png")), p,  width = 10, height = 5, dpi=600)
}
```

#### CC - for fraction separation
```{r VolcanoFraction}
#horizontal bar plots 
df<-ego_results_DE_CC_table
df<-df%>%
  filter(comparison %in% c("membrane - cytosol", "membrane_C - cytosol_C"))

df$logP <- -log10(df$p.adjust)
df <- df %>%
  arrange(desc(logP), .by_group = TRUE)%>%
  group_by(Cluster) %>%
  slice_max(order_by = logP, n = 10, with_ties = FALSE)

# df <- df %>% #if you want to annotate the genes contributing to each GO term
#   mutate(
#     geneID_wrapped = str_wrap(geneID, width = 50),  # wrap long gene lists
#     Description_label = paste0(Description, "\n", geneID_wrapped)  # Description on top, genes below
#   )

#define custom plot colours
df$direction <- factor(df$direction, levels = c("downregulated", "upregulated"))
direction_colors <- c("downregulated" = "#F8766D", "upregulated" = "#00B358")
df <- df %>%
  mutate(direction_colors = case_when(
    str_detect(tolower(Cluster), "_C") & direction == "downregulated" ~ "Downregulated control",
    str_detect(tolower(Cluster), "_C") & direction == "upregulated"   ~ "Upregulated control",
    !str_detect(tolower(Cluster), "_C") & direction == "downregulated" ~ "Downregulated patient",
    !str_detect(tolower(Cluster), "_C") & direction == "upregulated"   ~ "Upregulated patient",
    TRUE ~ "other" 
  ))
outline_color <- c("Downregulated control"   = "#F8766D",     
                    "Upregulated control"     = "#00B358",     
                    "Downregulated patient"  = "#F8766D",     
                    "Upregulated patient"    = "#00B358" )
fill_color <- c("Downregulated control"   = "white",     
                "Upregulated control"     = "white",       
                "Downregulated patient"  = "#F8766D",    
                "Upregulated patient"    = "#00B358")
for (s in unique(df$sample)) {
  plot_data <- df %>%      
  filter(sample == s)
          p <- ggplot(plot_data, aes(x = logP, y = reorder_within(Description, logP, Cluster), fill = direction_colors, color=direction_colors)) +
            geom_col(width = 0.4) +
            scale_fill_manual(values = fill_color) +
            scale_color_manual(values = outline_color) +
            scale_y_reordered() +
            facet_wrap(Cluster ~ ., scales = "free_y", space = "free_y") +
            theme_minimal() +
            labs(
              x = expression(-log[10](p.adjust)),
              y = NULL,
              title = "Cellular Compartment"
            ) +
            theme(
              panel.background = element_rect(fill = "white", color = NA), 
              plot.background = element_rect(fill = "white", color = NA),   
              axis.text.y = element_text(color = "black", size = 7, lineheight = 0.9),
              axis.text.x = element_text(color = "black"),
              plot.title = element_text(hjust = 0.5),
              strip.text.y = element_text(angle = 0, hjust = 0),
              panel.grid.major = element_blank(),  
              panel.grid.minor = element_blank(),
              strip.background = element_blank(),
              strip.text.x = element_blank(),
              legend.title = element_blank()
            )
          print(p)
          ggsave(file.path(dir_save, paste0("GO_enrichment_DE_CC_barplot_for fractions_", s, "_", script.version, "_", Sys.Date(), ".png")), p,  width = 6, height = 4, dpi=600)
}
```

### Molecular function
```{r GO_DE_MF}
ego_results_DE_MF <- NULL
try(ego_results_DE_MF <- compareCluster(
  ENTREZID ~ grouping + direction, #grouping by sample and up/downregulated
  data = ego_data,  #use dataset generated with ENTREZID
  fun = "enrichGO", #clusterProfiler::enrichGO()
  OrgDb = org.Hs.eg.db, #human Org database
  keyType = "ENTREZID",
  ont = "MF", #cellular compartment
  readable = TRUE,
  universe = ID_LUT$ENTREZID)) #background genes

if (!is.null(ego_results_DE_MF)) {
  ego_results_DE_MF_table <- ego_results_DE_MF@compareClusterResult %>% 
    mutate(comparison = gsub(".+___", "", grouping),
      sample = gsub("___.+", "", grouping)) %>%
    group_by(ID, sample, comparison, direction) %>%
    mutate(odds_ratio = calculateFE(GeneRatio, BgRatio)) %>%
    ungroup()
  
  ego_results_DE_MF_table$Description <- factor(
    ego_results_DE_MF_table$Description,
    levels = unique(rev(ego_results_DE_MF_table$Description)))

  ego_sub <- ego_results_DE_MF_table %>%
    mutate(is_cyt_mem = grepl("cytosol\\s*-\\s*cytosol_C", grouping, ignore.case = TRUE) |
                   grepl("membrane\\s*-\\s*membrane_C", grouping, ignore.case = TRUE)) %>%
    group_by(sample, comparison, direction) %>%
    arrange(p.adjust)%>% #arrange by smallest adjusted p-value first
    slice_head(n = 10) %>% # Keep top 10 per group
    group_by(is_cyt_mem) %>%
    arrange(p.adjust, .by_group = TRUE) %>%
    mutate(Description_reordered = factor(Description, levels = unique(Description))) %>%
    ungroup()
}

#horizontal bar plots 
df<-ego_results_DE_MF_table
df<-df%>%
  filter(comparison %in% c("membrane - membrane_C", "cytosol - cytosol_C"))

df$logP <- -log10(df$p.adjust)
df <- df %>%
  arrange(desc(logP), .by_group = TRUE)%>%
  group_by(Cluster) %>%
  slice_max(order_by = logP, n = 10, with_ties = FALSE)

#define custom plot colours
df$direction <- factor(df$direction, levels = c("downregulated", "upregulated"))
direction_colors <- c("downregulated" = "#F8766D", "upregulated" = "#00B358")
df <- df %>%
  mutate(direction_colors = case_when(
    str_detect(tolower(Cluster), "cytosol") & direction == "downregulated" ~ "Downregulated cytosol",
    str_detect(tolower(Cluster), "cytosol") & direction == "upregulated"   ~ "Upregulated cytosol",
    str_detect(tolower(Cluster), "membrane") & direction == "downregulated" ~ "Downregulated membrane",
    str_detect(tolower(Cluster), "membrane") & direction == "upregulated"   ~ "Upregulated membrane",
    TRUE ~ "other"))

ego_results_DE_MF_table2 <- ego_results_DE_MF_table %>%
  filter(comparison %in% c("cytosol - cytosol_C", "membrane - membrane_C")) %>%
  group_by(sample, comparison, direction) %>%
  arrange(p.adjust, .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  # Convert all factors to character before saving
  mutate(across(where(is.factor), as.character))

write.csv(ego_results_DE_MF_table2,
  file.path(dir_save, paste0("GO_enrichment_DE_MF_limma_results_top10_", script.version, ".csv")),
  row.names = FALSE)

#count how many times each gene appears in GO terms
gene_counts_MF <- ego_results_DE_MF_table %>%
  filter(comparison %in% c("cytosol - cytosol_C", "membrane - membrane_C"))%>%
  group_by(sample)%>%
  separate_rows(geneID, sep = "/") %>% # split by "/" 
  mutate(geneID = str_trim(geneID)) %>%
  count(geneID, sort = TRUE)%>% #count how many times each gene appears
  pivot_wider(
    names_from = sample,
    values_from = n,
    values_fill = 0 )

write.csv(gene_counts_MF, file.path(dir_save, paste0("GO_gene_counts_MF_", script.version, ".csv")), row.names = FALSE)

#define custom plot colours
df$direction <- factor(df$direction, levels = c("downregulated", "upregulated"))
direction_colors <- c("downregulated" = "#F8766D", "upregulated" = "#00B358")
df <- df %>%
  mutate(direction_colors = case_when(
    sample == "CTX" & direction == "downregulated" ~ "Downregulated CTX",
    sample == "CTX"  & direction == "upregulated"   ~ "Upregulated CTX",
    sample == "INT"  & direction == "downregulated" ~ "Downregulated INT",
    sample == "INT"  & direction == "upregulated"   ~ "Upregulated INT",
    TRUE ~ "other" 
  ))
outline_color <- c("Downregulated CTX"   = "#F8766D",     
                    "Upregulated CTX"     = "#00B358",     
                    "Downregulated INT"  = "#F8766D",     
                    "Upregulated INT"    = "#00B358" )
fill_color <- c("Downregulated CTX"   = "#F8766D",     
                "Upregulated CTX"     = "#00B358",       
                "Downregulated INT"  = "white",    
                "Upregulated INT"    = "white")
df$direction_colors <- factor(df$direction_colors, levels = c("Downregulated CTX", "Upregulated CTX", "Downregulated INT", "Upregulated INT"))

#plot all together - separated by cytosol/membrane
for (s in unique(df$comparison)) {
  plot_data <- df %>%      
  filter(comparison == s)#%>%
    #mutate(Description = str_wrap(Description, width = 50))

  
p <- ggplot(plot_data, aes(x = logP, y = reorder_within(Description, logP, Cluster), fill = direction_colors, color=direction_colors)) +
            geom_col(width = 0.4) +
            scale_fill_manual(values = fill_color) +
            scale_color_manual(values = outline_color) +
            scale_y_reordered() +
            facet_wrap(Cluster ~ ., scales = "free", space = "free_y") +
            theme_minimal() +
            labs(x = expression(-log[10](p.adjust)),
              y = NULL, title = "ii) Molecular Function") +
            theme(panel.background = element_rect(fill = "white", color = NA), 
              plot.background = element_rect(fill = "white", color = NA),   
              axis.text.y = element_text(color = "black", size = 7, lineheight = 0.9),
              axis.text.x = element_text(color = "black"),
              plot.title = element_text(hjust = 0.5),
              strip.text.y = element_text(angle = 0, hjust = 0),
              panel.grid.major = element_blank(),  
              panel.grid.minor = element_blank(),
              strip.background = element_blank(),
              strip.text.x = element_blank(),
              legend.title = element_blank())+
          theme(legend.position = "none")
          print(p)
          ggsave(file.path(dir_save, paste0("GO_enrichment_DE_MF_barplot_", s, "-",  script.version, "_", Sys.Date(), ".png")), p,  width = 5.05, height = 3.2, dpi=600)
}
```

### Biological process

```{r GO_DE_BP}
ego_results_DE_BP <- NULL
try(ego_results_DE_BP <- compareCluster(
  ENTREZID ~ grouping + direction, #grouping by sample and up/downregulated
  data = ego_data,  #use dataset generated with ENTREZID
  fun = "enrichGO", #clusterProfiler::enrichGO()
  OrgDb = org.Hs.eg.db, #human Org database
  keyType = "ENTREZID",
  ont = "BP", #cellular compartment
  readable = TRUE,
  universe = ID_LUT$ENTREZID)) #background genes

if (!is.null(ego_results_DE_BP)) {
  ego_results_DE_BP_table <- ego_results_DE_BP@compareClusterResult %>% 
    mutate(comparison = gsub(".+___", "", grouping),
      sample = gsub("___.+", "", grouping)) %>%
    group_by(ID, sample, comparison, direction) %>%
    mutate(odds_ratio = calculateFE(GeneRatio, BgRatio)) %>%
    ungroup()
  
  ego_results_DE_BP_table$Description <- factor(
    ego_results_DE_BP_table$Description,
    levels = unique(rev(ego_results_DE_BP_table$Description)))

  ego_sub <- ego_results_DE_BP_table %>%
    mutate(is_cyt_mem = grepl("cytosol\\s*-\\s*cytosol_C", grouping, ignore.case = TRUE) |
                   grepl("membrane\\s*-\\s*membrane_C", grouping, ignore.case = TRUE)) %>%
    group_by(sample, comparison, direction) %>%
    arrange(p.adjust)%>% #arrange by smallest adjusted p-value first
    slice_head(n = 10) %>% # Keep top 10 per group
    group_by(is_cyt_mem) %>%
    arrange(p.adjust, .by_group = TRUE) %>%
    mutate(Description_reordered = factor(Description, levels = unique(Description))) %>%
    ungroup()
}

#horizontal bar plots 
df<-ego_results_DE_BP_table
df<-df%>%
  filter(comparison %in% c("membrane - membrane_C", "cytosol - cytosol_C"))

df$logP <- -log10(df$p.adjust)
df <- df %>%
  arrange(desc(logP), .by_group = TRUE)%>%
  group_by(Cluster) %>%
  slice_max(order_by = logP, n = 10, with_ties = FALSE)

#define custom plot colours
df$direction <- factor(df$direction, levels = c("downregulated", "upregulated"))
direction_colors <- c("downregulated" = "#F8766D", "upregulated" = "#00B358")
df <- df %>%
  mutate(direction_colors = case_when(
    str_detect(tolower(Cluster), "cytosol") & direction == "downregulated" ~ "Downregulated cytosol",
    str_detect(tolower(Cluster), "cytosol") & direction == "upregulated"   ~ "Upregulated cytosol",
    str_detect(tolower(Cluster), "membrane") & direction == "downregulated" ~ "Downregulated membrane",
    str_detect(tolower(Cluster), "membrane") & direction == "upregulated"   ~ "Upregulated membrane",
    TRUE ~ "other"))

ego_results_DE_BP_table2 <- ego_results_DE_BP_table %>%
  filter(comparison %in% c("cytosol - cytosol_C", "membrane - membrane_C")) %>%
  group_by(sample, comparison, direction) %>%
  arrange(p.adjust, .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  # Convert all factors to character before saving
  mutate(across(where(is.factor), as.character))

write.csv(ego_results_DE_BP_table2,
  file.path(dir_save, paste0("GO_enrichment_DE_BP_limma_results_top10_", script.version, ".csv")),
  row.names = FALSE)

#count how many times each gene appears in GO terms
gene_counts_BP <- ego_results_DE_BP_table %>%
  filter(comparison %in% c("cytosol - cytosol_C", "membrane - membrane_C"))%>%
  group_by(sample)%>%
  separate_rows(geneID, sep = "/") %>% # split by "/" 
  mutate(geneID = str_trim(geneID)) %>%
  count(geneID, sort = TRUE)%>% #count how many times each gene appears
  pivot_wider(
    names_from = sample,
    values_from = n,
    values_fill = 0 )

write.csv(gene_counts_BP, file.path(dir_save, paste0("GO_gene_counts_BP_", script.version, ".csv")), row.names = FALSE)

#define custom plot colours
df$direction <- factor(df$direction, levels = c("downregulated", "upregulated"))
direction_colors <- c("downregulated" = "#F8766D", "upregulated" = "#00B358")
df <- df %>%
  mutate(direction_colors = case_when(
    sample == "CTX" & direction == "downregulated" ~ "Downregulated CTX",
    sample == "CTX"  & direction == "upregulated"   ~ "Upregulated CTX",
    sample == "INT"  & direction == "downregulated" ~ "Downregulated INT",
    sample == "INT"  & direction == "upregulated"   ~ "Upregulated INT",
    TRUE ~ "other" 
  ))
outline_color <- c("Downregulated CTX"   = "#F8766D",     
                    "Upregulated CTX"     = "#00B358",     
                    "Downregulated INT"  = "#F8766D",     
                    "Upregulated INT"    = "#00B358" )
fill_color <- c("Downregulated CTX"   = "#F8766D",     
                "Upregulated CTX"     = "#00B358",       
                "Downregulated INT"  = "white",    
                "Upregulated INT"    = "white")
df$direction_colors <- factor(df$direction_colors, levels = c("Downregulated CTX", "Upregulated CTX", "Downregulated INT", "Upregulated INT"))

#plot all together - separated by cytosol/membrane
for (s in unique(df$comparison)) {
  plot_data <- df %>%      
  filter(comparison == s)%>%
    mutate(Description = str_wrap(Description, width = 50))
  
p <- ggplot(plot_data, aes(x = logP, y = reorder_within(Description, logP, Cluster), fill = direction_colors, color=direction_colors)) +
            geom_col(width = 0.4) +
            scale_fill_manual(values = fill_color) +
            scale_color_manual(values = outline_color) +
            scale_y_reordered() +
            facet_wrap(Cluster ~ ., scales = "free", space = "free_y") +
            theme_minimal() +
            labs(x = expression(-log[10](p.adjust)),
              y = NULL, title = "iii) Molecular Function") +
            theme(panel.background = element_rect(fill = "white", color = NA), 
              plot.background = element_rect(fill = "white", color = NA),   
              axis.text.y = element_text(color = "black", size = 7, lineheight = 0.9),
              axis.text.x = element_text(color = "black"),
              plot.title = element_text(hjust = 0.5),
              strip.text.y = element_text(angle = 0, hjust = 0),
              panel.grid.major = element_blank(),  
              panel.grid.minor = element_blank(),
              strip.background = element_blank(),
              strip.text.x = element_blank(),
              legend.title = element_blank())+
          theme(legend.position = "none")
          print(p)
          ggsave(file.path(dir_save, paste0("GO_enrichment_DE_BP_barplot_", s, "-",  script.version, "_", Sys.Date(), ".png")), p,  width = 3.5, height = 4, dpi=600)
}
```

# Save Methods for analysis

```{r SaveMethods}
method.description <- "For the proteomics data analysis the raw output files of FragPipe (protein.tsv files files) were processed using the R programming environment (ISBN 3-900051-07-0)."
method.description <- append(method.description, "Initial data processing included filtering out contaminants and reverse proteins. Only proteins quantified with at least 2 razor peptides (with Razor.Peptides >= 2) were considered for further analysis.")
method.description <- append(method.description, paste(nrow(cdata), "proteins passed the quality control filters."))
method.description <- append(method.description, glue("In order to correct for technical variability, batch effects were removed using the 'removeBatchEffect' function of the limma package (PMID: 25605792) on the log2 transformed raw TMT reporter ion intensities ('channel' columns). Subsequently, normalization was performed using the 'normalizeVSN' function of the limma package (VSN - variance stabilization normalization - PMID: 12169536)."))
method.description <- append(method.description, "Differential expression analysis was performed using the moderated t-test provided by the limma package (PMID: 25605792). The model accounted for replicate information by including it as a factor in the design matrix passed to the 'lmFit' function.")
if ("fdrtool" %in% limma_results$hit_annotation_method)
{
  method.description <- append(method.description, "To obtain p-values and false discovery rates (FDRs), the 'fdrtool' function from the fdrtool package (PMID: 18441000) was used to analyze the t-values produced by limma for certain comparisons.")
}
method.description <- append(method.description, glue("Proteins were annotated as hits if they had a false discovery rate (FDR) below {fdr_hit_threshold} and an absolute fold change greater than {fc_hit_threshold}." ))
method.description <- append(method.description, glue("Clustering {heatmap_selection_label} based on the median protein abundances normalized by median of control condition was conducted to identify groups of proteinsimilar patterns across conditions. The 'kmeans' method was employed, using Euclidean distance as the distance metric and 'ward.D2' linkage for hierarchical clustering. The optimal number of clusters ({cluster.number}) was determined using the Elbow method, which identifies the point where the within-group sum of squares stabilizes."))
method.description <- append(method.description, "Gene ontology (GO) enrichment analysis was performed using the 'compareCluster' function of the 'clusterProfiler' package (PMID: 22455463), which assesses over-representation of GO terms in the dataset relative to the background gene set. Enrichment was conducted for the following GO categories: Cellular Component (CC), Molecular Function (MF), and Biological Process (BP). The analysis was performed using 'org.Hs.eg.db' as the reference database. The odds ratio ('odds_ratio') for each GO term was calculated by comparing the proportion of genes associated with that term in the dataset ('GeneRatio') to the proportion in the background set ('BgRatio'). An odds ratio greater than 1 indicates that the GO term is enriched in the dataset compared to the expected background.")
method.description <- paste(method.description, collapse = " ")
writeLines(text = method.description, con = file.path(dir_save, paste0("Methods_", script.version, ".txt")))
```

# Save R workspace

```{r}
#save.image(file.path(dir_save, paste0("Workspace_", script.version, ".RData")))
# load(file.path(dir_save, paste0("Workspace_", script.version, ".RData")))
```

# Save Session information

```{r SaveSessionInfo}
sessionInfo()
```

